---
title: "R Basics"
output:
  html_document:
    toc: yes
Objectives: 
- "To ensure all participants have the tools to participate in this workshop"
---



# Software


R is a programming language that is especially powerful for data exploration, visualization, and statistical analysis. To interact with R, we will use RStudio.

The following software is required for this workshop:

### R [Free]: The R Project for Statistical Computing.

- Download R for your platform here: https://cloud.r-project.org/
- Install

### RStudio [Free]: A powerful and productive user interface for R. It's free and open source, and works great on Windows, Mac, and Linux.

- https://www.rstudio.com/products/rstudio/download/
- Download and install
- Attention: Download and install R FIRST! from the link above.


#### Attention Mac users!

Some functions in R require an "X11 Server" and/or libraries associated with an X11 server. Apple does not provide this software with OS X anymore so you have to do it on your own via a third-party application called XQuartz.

1. Please close all the programs.
2. Go to http://xquartz.macosforge.org/landing/ and download the top-most "quick download" disk image file.
3. Double click the downloaded disk image then double click on the XQuartz.pkg and follow the installation
steps.
4. Please restart your machine.

# R Basics

R is a powerful language and environment for statistical computing. R is a freeware and can be installed on any computers for free. R is frequently used to analyze and visualize data in many scientific applications. R in combination with RStudio is a great option for educators as it allows easy data simulations, data analysis, and data visualization all within one powerful interface. This short tutorial will be covering basics of R use within RStudio interface. 

## RStudio layout

**Bottom left:** console window for typing simple commands.

**Top left:** editor window. Collection of commands (scripts) can be edited here using a simple word editor.

**Top right:** workspace/history window. Here you can view data and values loaded into R memory.

**Bottom right:** Files/plots/packages/help window. Here you can navigate and open your files, see your plots, load necessary package, and search for help.

## Getting help

1. The most productive way to get help, by far, is just to google your question. R has a thriving and vibrant community, it is almost certain that the answer is out there.

2. "A (very) short introduction to R" reading can be found at the end of this document (please refer to table of contents). This very short article describes all the basics of R and is a very helpful guide for those who need help with the material we cover in class.

3. The CRAN Introduction to R provides a more complete and detailed overview of the entire R language [http://cran.r-project.org/doc/manuals/R-intro.pdf]

4. Quick-R website has a great collection of articles for novice to advanced R users [http://www.statmethods.net/].

5. Once R is installed, there is a comprehensive built-in help system. At the program's command prompt you can use any of the following:

- help.start() # general help
- help(foo) # help about function foo
- ?foo # same thing
- apropos("foo") # list all functions containing string foo
- example(foo) # show an example of function foo
- vignette() # show available vignettes
- vignette("foo")# show specific vignette


## Creating and executing scripts

- Open RStudio
- Create new R script file: File ! New File ! R Script
- Save this script file in our Project folder (name it RScript)

In the editor window (newly created R script file) let's type:

```{r, eval=FALSE}

print("Hello world!")
```

Make sure that cursor is on the line with print("Hello world") and hit Ctrl+Enter 

R is a compiled or interpreted language. Your command will be immediately executed and you should see:

```{r, echo=FALSE}

print("Hello world!")
```


in the console window.


To execute a piece of code you can highlight that code and hit *Run* on the top right of your editor window. You can also place a cursor on the line of code you want to execute and hit *Ctrl+Enter* or hit Run at the top of your editor window.

## Using R as a calculator

In the command window after the ">" type:

```{r, eval=FALSE}
2+2
```

when you hit Enter you should see this output


```{r, echo=FALSE}
2+2
```

## Assignment operator

"<-" used to assign a value to a variable

### Example 1:

```{r, eval=FALSE}
cat<-2
dog<-3
cat*dog

```

when you hit Enter you should see this output

```{r, echo=FALSE}
cat<-2
dog<-3
cat*dog

```


What we did is:

- we assigned 2 to variable "cat"
- we assigned 3 to variable "dog"
- we then multiplied those two variables

### Example 2:

```{r}
x <- c(1,2,3,4,5,6)
y<-x^2
print(y)
```



In the code above, "c" is a function that combines its arguments into a vector (a string of numbers).




















# Session details

```{r}
devtools::session_info()
```
