---
title: "Project Management"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

# Creating projects in RStudio

> File: New Project

OR

> Project list (top right corner): New Project

## Project structure

I recommend 

* two folders in the root of your project directory: figures-tables and data
* ProjectName.Rmd file for data visualization and data analyses

Note! RStudio project files do not play well with _Box Drive_. They do work well with _Box Sync_ and _OneDrive_ (I have not tested dropbox).


### Project tree structure

* MainProjectFolder/
    + data/
    + figures-tables/
    + ProjectName.Rmd
    + ProjectName.Rproj
        


## Data - Practice dataset

1. Go to <a href="https://github.com/sergioschr/R_workshop/blob/master/data/Sample-Dose-Effect-Curve.csv" target="_blank">GitHub</a>

2. Right click on "Raw"

3. "Save link as" to save it in project/data folder


## Code files

Let's create a working file for your data visualization, figures, and data analysis.


### Initial setup

1. Create Rmd file -> File:New File:R Markdown:PDF output
2. Erase body (everything below preamble)
3. Insert new R chunk (see +C "Insert" in the document top toolbar)
4. Paste the following code in the first chunk

```{r message=FALSE, warning=FALSE, include=FALSE}

ipak <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg))
                install.packages(new.pkg, dependencies = TRUE)
        sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("ggplot2", "Rmisc", "tidyr", "RColorBrewer", "plotly", "ggplot2", "nlme", "dplyr", "grid", "gridExtra") # add or remove packages as needed
ipak(packages)
```

5. Adjust chunk parameters

> {r, message=FALSE, warning=FALSE, include=FALSE}

* message=FALSE # turn off all messages
* warning=FALSE # turn off all warnings
* include=FALSE # do not show output
* find more options by googling "rstudio chunk options"
* you can adjust these parameters by using gear icon on the top right of each code chunk. 

### Add project description using markdown (markdown cheat sheet can be found in RStudio under Help:Cheatsheets)

> \# My ground braking experiment

> \#\# Overview

> This project investigates the effect of "x" on "y" in both male and female subjects.


### Import and assess data

* Insert another R chunk with the following code


```{r}
dataset <- read.csv("data/Sample-Dose-Effect-Curve.csv")
summary(dataset)

```

* Insert a final chunk with heading \#\#\# Data visualization

```{r}
ggplot(dataset, aes(x = factor(IV.DOSE), y = DV.Response, colour=IV.Sex)) +
        geom_boxplot()
```


* The plot is presented in a default ggplot configuration. This configuration is not often desirable for the publication.
* Let's adjust some parameters to make it a bit more publication friendly. 

### Convert variables to factors and rename




```{r}
dataset <- read.csv("data/Sample-Dose-Effect-Curve.csv")
dataset$Subject<-as.factor(dataset$Subject)
dataset$Drug<-as.factor(dataset$Drug)
dataset$Sex <-as.factor(dataset$IV.Sex )
dataset$Dose <- factor(dataset$IV.DOSE, levels=c("0","5","10", "20", "30", "60"))
summary(dataset)

# See RStudio Help->Cheatsheets for more information about data transformations, manipulations, and visualization. 

# Classic theme

ggplot(dataset, aes(x = Dose, y = DV.Response, colour=Sex)) +
        geom_boxplot()+
        theme_classic()+
        labs(title = "Title of your plot", subtitle = "Subtitle of your plot", y="Renamed y axis")


# bw theme

ggplot(dataset, aes(x = Dose, y = DV.Response, colour=Sex)) +
        geom_boxplot()+
        theme_bw()+
        labs(title = "Title of your plot", subtitle = "Subtitle of your plot", y="Renamed y axis")
```
For more information on different ggplot themes see

https://ggplot2-book.org/polishing.html
https://ggplot2.tidyverse.org/reference/ggtheme.html


### Create a PDF output

* Just click "Knit" on the document's top toolbar.


### Additional learning opportunity

* place your own dataset in the data folder
* create separate Rmd data visualization file with distinct name of your dataset
* within the Rmd file create code to summarize your data and visualize your data as you like or able.
* try different themes for your plot
* try different Titles for various elements in your plot
* output Rmd to PDF file using "Knit" function
* Troubleshoot if needed by googling your problem.
* Find more information here https://ggplot2-book.org/


# Session details




```{r}
devtools::session_info()
```


# <a href="/R_workshop" class="topic" style="color:red">Navigate Back</a>