---
title: "Project Management"
output:
  html_document:
    toc: yes
---

# Creating projects in RStudio

> File: New Project

OR

> Project list (top right corner): New Project

## Project structure

I recommend 

* two additional folders: figures and data
* data-visualization.Rmd file for data visualization
* figures.R file for publication quality figures
* data-analysis.R for all your data analysis

### Project treee structure

* MainProjectFolder/
    + data/
    + figures/
    + data-analysis.R
    + data-visualization.Rmd
    + figures.R 
    + ProjectName.Rproj
        


## Data - Practice dataset

1. Go to <a href="https://github.com/sergioschr/R_workshop/blob/master/data/Sample-Dose-Effect-Curve.csv" target="_blank">GitHub</a>

2. Right click on "Raw"

3. "Save link as" to save it in project/data folder


## Code files

Let's create data visualization, figures, and data analysis files in your project directory

## Data visualization file

Let's start with data-visualization.Rmd file

### Initial setup

1. Create file File:New File:R Markdown:PDF output
2. Erase body (everything below preamble)
3. Insert new R chunk (see +C "Insert" in the document top toolbar)
4. Paste the following code in the first chunk

```{r, message=FALSE, warning=FALSE}

ipak <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg))
                install.packages(new.pkg, dependencies = TRUE)
        sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("ggplot2", "Rmisc", "tidyr", "RColorBrewer", "plotly", "ggplot2", "nlme", "dplyr", "cowplot", "grid", "gridExtra") # add or remove packages as needed
ipak(packages)
```

5. Adjust chunk parameters

> {r, message=FALSE, warning=FALSE}

* message=FALSE # turn off all messages
* warning=FALSE # turn off all warnings
* find more options by googling "rstudio chunk options"

### Add project description using markdown (markdow cheat sheet can be found in RStudio under Help:Cheatsheets)

> \# My ground braking experiment

> \#\# Overview

> This project investigates the effect of "x" on "y" in both male and female subjects.


### Import and assess data

* Insert another R chunk with the following code


```{r}
dataset <- read.csv("data/Sample-Dose-Effect-Curve.csv")
summary(dataset)

```

* Insert a final chunk with heading \#\#\# Data visualization

```{r}
ggplot(dataset, aes(x = factor(IV.DOSE), y = DV.Response, colour=IV.Sex)) +
        geom_boxplot()
```


* Note that plot now looks different that before - no grid lines, white background, and no top and right border thanks to "cowplot" package. Once "cowplot" package is loaded ggplot plots will be automatically cleaned up. 



### Create a PDF output

* Just click "Knit" on the document's top toolbar.


### Additional learning opportunity

* place your own dataset in the data folder
* create separate data visualization file with distinct name of your dataset
* within the Rmd file create code to summarize your data and visualize your data as you like or able.
* output Rmd to PDF file using "Knit" function
* Troubleshoot if needed by googling your problem.




# Session details




```{r}
devtools::session_info()
```


# <a href="/R_workshop" class="topic" style="color:red">Navigate Back</a>