---
title: "Basics of Data Analysis using R"
output:
  html_document:
    toc: yes
---


# Basics of Statistical Analysis Using R

## Power analysis

To perform power analysis we will be using `pwr` package. Please installed this package using `install.packages()` function.

```{r}
library (pwr) # load required package

# Run power analysis to estimate sample size (n)

## type="paired" indicates that experiment used the same subjects in both conditions (within-subjects design). Note that "d" is set to very low effect size of 0.2.

pwr.t.test(n = , d = 0.2, sig.level = 0.05 , power = 0.80, type = "paired")

## type="two.sample" indicates that experiment used different groups of subjects in two conditions (between-subjects design). Note that "d" is set to very low effect size of 0.2.

pwr.t.test(n = , d = 0.2, sig.level = 0.05 , power = 0.80, type = "two.sample")


#Plot sample size - power relationship

plot(pwr.t.test(d = 0.2, power = 0.80, type = "two.sample"))

# Now let's visualize paired design with good effect size

plot(pwr.t.test(d = 0.8, power = 0.80, type = "paired"))

```




## T-test

```{r}

# Simulate data for t-test

before<-(rnorm(14, mean=22, sd=10))
after<-(rnorm(14, mean=25, sd=10))

boxplot(before, after)

# Analyze data using t-test

t.test(before, after, paired=TRUE) # paired. Is the difference between these groups significant?


# Change data parameters

before<-(rnorm(14, mean=22, sd=1))
after<-(rnorm(14, mean=25, sd=1))

boxplot(before, after)

t.test(before, after, paired=TRUE) # Is the difference between these groups significant? What changed?


```






## Analysis of variance

### Simulate dataset

```{r}
### Females ###

# simulate level 1 for females
subject<-(1:10)
sex<-replicate(10, "female")
level<-replicate(10, "control")
dv<-rnorm(10, mean=10, sd=1)
dff1<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dff1

# simulate level 2 for females
subject<-(11:20)
sex<-replicate(10, "female")
level<-replicate(10, "low")
dv<-rnorm(10, mean=12, sd=2)
dff2<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dff2

# simulate level 3 for females
subject<-(21:30)
sex<-replicate(10, "female")
level<-replicate(10, "high")
dv<-rnorm(10, mean=18, sd=2.5)
dff3<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dff3


### Males ###

# simulate level 1 for males
subject<-(31:40)
sex<-replicate(10, "male")
level<-replicate(10, "control")
dv<-rnorm(10, mean=18, sd=1.2)
dfm1<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dfm1

# simulate level 2 for males
subject<-(41:50)
sex<-replicate(10, "male")
level<-replicate(10, "low")
dv<-rnorm(10, mean=15, sd=1.8)
dfm2<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dfm2

# simulate level 3 for males
subject<-(51:60)
sex<-replicate(10, "male")
level<-replicate(10, "high")
dv<-rnorm(10, mean=11, sd=2.1)
dfm3<-data.frame(subject=subject, sex=sex, dose=level, dv=dv)
dfm3

# merge data frames into one dataset

df<-rbind(dff1, dff2, dff3, dfm1, dfm2, dfm3)
df
str(df)


```

### Visualize data and run simple ANOVAs

```{r}
# Visualize dataset

library(ggplot2)

ggplot(df, aes(x=dose, y=dv, colour=sex)) +
        geom_boxplot()+
        facet_wrap(~sex)

library(Rmisc)

dfse<-summarySE(df, measurevar="dv", groupvars=c("sex", "dose"))


ggplot(dfse, aes(x=dose, y=dv, group=sex, colour=sex)) +
        geom_line() +
        geom_point()+
        geom_errorbar(aes(ymin=dv-se, ymax=dv+se), width=.1)


## Extract female data

female<-subset(df, sex=="female")
female

## Extract male data

male<-subset(df, sex=="male")
male

## Omnibus ANOVA

summary(aov(dv~dose, df))

## One way ANOVA - Females

summary(aov(dv~dose, female))
TukeyHSD(aov(dv~dose, female)) # "post hoc" multiple comparisons

## One way ANOVA - Males

summary(aov(dv~dose, male))
TukeyHSD(aov(dv~dose, male)) # "post hoc" multiple comparisons

## Two way ANOVA - Sex X Dose

summary(aov(dv~sex*dose, df))
TukeyHSD(aov(dv~sex*dose, df)) # "post hoc" multiple comparisons

```




### ANOVA on a more complex dataset

```{r, message=FALSE, warning=FALSE}
ipak <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg))
                install.packages(new.pkg, dependencies = TRUE)
        sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("ggplot2", "Rmisc", "tidyr", "RColorBrewer", "plotly", "ggplot2", "nlme", "dplyr", "cowplot", "grid", "gridExtra", "reshape", "plyr") # add or remove packages as needed
ipak(packages)
```


```{r}
# import data
mydataset <- read.csv("data/Sample-Dose-Effect-Curve.csv")

#Explore dataset
summary(mydataset)
str(mydataset)

# Rename columns

mydataset<-rename(mydataset, c("IV.Sex"="Sex", "DV.Activity"="Activity", "IV.DOSE"="DOSE", "DV.Response"="Response"))
str(mydataset)

# Convert variables to factors

mydataset$Subject<-as.factor(mydataset$Subject)
mydataset$DOSE<-as.factor(mydataset$DOSE)

# Rename level of a factor

with(mydataset, (levels(Sex)))
mydataset$Sex<-with(mydataset, revalue(Sex, c("Fem"="F")))
str(mydataset)

### Subset dataset for separate analysis if needed
Female<-subset(mydataset, Sex=="F")
Male<-subset(mydataset, Sex=="M")

summary(Female)
summary(Male)


#Basic summary statistics by group usign Psych package
library(psych)
with(mydataset, (describeBy(Response, list(DOSE, Sex), mat=TRUE)))

# Preliminary visualization

#Boxplot
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot()

# save boxplot as bp1
bp1<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot()


#Boxplot with individual data
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot() +
geom_jitter()

# save as bp2
bp2<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot() +
  geom_jitter()


#Boxplot: by Sex
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot(aes(fill=factor(Sex)))

# save as bp3

bp3<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot(aes(fill=factor(Sex)))


#Line plot: DOSE by Sex

ggplot(mydataset, aes(x=DOSE, y=Response, colour=Sex, group=Sex)) + 
  geom_smooth()


```

#### Considerations

* Type I, II, and III can be modeled using "ez" package and a bit of googling or textbook referral
* By default, R uses Type I sums of squares, and SPSS uses Type III sums of squares
* There is a debate out there about what is more appropriate depending on conditions
* Certain assumptions have to be met (homogeneity of variance)
* Distributions within groups should be normally distributed
* General format: aov.model <-aov(outcome ~ predictor(s) + Error, data=dataFrame)
* Error term: Error(Subject/Within Subjects IV) or Error if one between subjects variable - Error(Subject)
* Suggested resourse: http://egret.psychol.cam.ac.uk/statistics/R/anova.html#visualization

```{r}
# Two way mixed factor ANOvA
AnovaModel.1 <- aov(Response ~ Sex*DOSE + Error(Subject/DOSE), data=mydataset)
summary(AnovaModel.1)

# One way Anova with Sex removed
AnovaModel.2 <- aov(Response ~ DOSE + Error(Subject/DOSE), data=mydataset)
summary(AnovaModel.2)


# One way Female only
AnovaModel.3 <- aov(Response ~ DOSE + Error(Subject), data=Female)
summary(AnovaModel.3)


# One way Male only
AnovaModel.4 <- aov(Response ~ DOSE + Error(Subject), data=Male)
summary(AnovaModel.4)
```



# <a href="/R_workshop" class="topic" style="color:red">Navigate Back</a>