---
title: "Basics of Data Analysis using R (Part 2)"
output:
  html_document:
    toc: yes
---

## Load packages
```{r, message=FALSE, warning=FALSE}

ipak <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg))
                install.packages(new.pkg, dependencies = TRUE)
        sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("ggplot2", "Rmisc", "tidyr", "RColorBrewer", "plotly", "ggplot2", "nlme", "dplyr", "cowplot", "grid", "gridExtra", "reshape", "plyr", "psych") # add or remove packages as needed
ipak(packages)
```

#Importing and cleaning data

##Import data

```{r}
#Load dataset
mydataset <- read.csv("data/Sample-Dose-Effect-Curve.csv")

#Explore dataset
summary(mydataset)
str(mydataset)

```

##Rename columns

```{r}


mydataset<-rename(mydataset, c("IV.Sex"="Sex", "DV.Activity"="Activity", "IV.DOSE"="DOSE", "DV.Response"="Response"))
str(mydataset)

```

##Convert variables to factors

```{r}
mydataset$Subject<-as.factor(mydataset$Subject)
mydataset$DOSE<-as.factor(mydataset$DOSE)

#Explore dataset
summary(mydataset)
str(mydataset)
```


## Rename level of a factor

```{r}
with(mydataset, (levels(Sex)))
mydataset$Sex<-with(mydataset, revalue(Sex, c("Fem"="F")))
str(mydataset)

```


##Subset 

```{r}
### Subset dataset for separate analysis if needed
Female<-subset(mydataset, Sex=="F")
Male<-subset(mydataset, Sex=="M")

summary(Female)
summary(Male)
```


##Basic summary statistics using Psych package

```{r}
#Basic summary statistics by group
with(mydataset, (describeBy(Response, list(DOSE, Sex), mat=TRUE)))

# How about a nicer view of this table
library(knitr) # use knitr package

# use kable function
kable(with(mydataset, (describeBy(Response, list(DOSE, Sex), mat=TRUE))))
```

##Preliminary visualization 

```{r}
#Boxplot
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot()

# save boxplot as bp1
bp1<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot()


#Boxplot with individual data
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot() +
geom_jitter()

# save as bp2
bp2<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot() +
  geom_jitter()


#Boxplot: by Sex
p<-ggplot(mydataset, aes(factor(DOSE), Response))
p+geom_boxplot(aes(fill=factor(Sex)))

# save as bp3

bp3<-ggplot(mydataset, aes(factor(DOSE), Response))+
  geom_boxplot(aes(fill=factor(Sex)))


#Line plot: DOSE by Sex

ggplot(mydataset, aes(x=DOSE, y=Response, colour=Sex, group=Sex)) + 
  geom_smooth()
  
  
  
```


# Multiple (Linear) Regression



```{r}
# Comparing models
fit0<-lm(Response~1, data=mydataset)
fit1<-lm(Response~DOSE, data=mydataset)
fit2<-lm(Response~DOSE+Sex, data=mydataset)
anova(fit0,fit1,fit2)

# Fitting model
fit <- lm(Response~DOSE+Sex, data=mydataset)
anova(fit) # anova table 
summary(fit)

fit <- lm(Response~DOSE*Sex, data=mydataset)
anova(fit) # anova table 
summary(fit)

# Other useful functions 
coefficients(fit) # model coefficients
confint(fit, level=0.95) # CIs for model parameters 
vcov(fit) # covariance matrix for model parameters 
plot(fit)

# diagnostic plots 
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(fit)

```




# Basics of Statistical Analysis Using R (Part 2)

## Multilevel approach: multilevel linear model

* See p. 573 of Discovering Statistics Using R, 1st Edition by Andy Field, Jeremy Miles, Zoe Field.
* Well suited to repeated-measures experimental designs
* Can be modeled using _lme()_ function from {nlme} package
* General format: lme.model<-lme(outcome ~ predictor(s), random = random effects, data = dataFrame, method="ML")
* The default method is restricted maximum-likelihood estimation (REML)
* For various reasons I am not competent enough to discuss here, the advised method is maximum likelihood (ML)
* "random" specifies any random effects - an effect that can vary across different entities
* In the example below we have _random=~1 | Subject/DOSE/Sex_ which means that the data with the same value of _Subject_ within different levels of _DOSE_ or _Sex_ are dependent (i.e., from the same subject)

### Step by Step: Using multilevel approach

1. Build a model from a baseline that includes no predictors other than intercept.
2. Add predictor 1 (DOSE)
3. Add predictor 2 (Sex)
4. Add interaction (Dose:Sex)
5. Compare models to the baseline using anova
6. Determine if adding predictors and/or interaction significantly improved the model**
7. Run appropriate "summary" on the model of interest which will give you all the effects and comparisons***

** When added predictor significantly improves a model, in other words the predictor had a significant effect on dependent measure, we can record this as such:

$$\chi^2(2)=9.1, p=.010$$

Where (2) is an increase of degrees of freedom from baseline, in this case by 2

*** When reporting contrasts from "summary" output we can print it as such:

$$b=2.19, t(38)=3.18, p=.003$$

See Andy Field pages 595-596


```{r}

# Build a baseline: no predictors other than the intercept

baseline<-lme(Response~1, random=~1 | Subject/DOSE/Sex, data=mydataset, method="ML")

# note that the outcome is predicted only from the intercept (Response~1)
# and we asked to use ML to estimate the model
# the variables DOSE/Sex are nested within the variable Subject (scores for levels of thes variables can be found within each Sbuject)
# to see the effect of each predictor we need to add them one by one
# lets add "DOSE" variable to the model

doseModel<-update(baseline, .~. +DOSE)

# note that we created new model "doseModel" so we can compare it to "baseline" above if need it
# note that we used update function to simply update the model above without rewriting most of it
# next, we can keep buildign the modely by adding "Sex" variable to the "doseModel"


sexModel<-update(doseModel, .~. +Sex)

# note that we created a new model "sexModel" using "update" function
# next, lets add interaction term

mixModel<-update(sexModel, .~. +DOSE:Sex)

# to compare these models we can list them in the order in which we want them to be compared using anova() function

anova(baseline, doseModel, sexModel, mixModel)

# We can further explore the model by runnign summary function

summary(doseModel) 

# this give us all the comparisons to 0 we need!
# next lets explore sexModel

summary(sexModel)

# looks like there "Sex" did not significantly contribued to the variance
# finally lets explore mixModel

summary(mixModel)


```


# <a href="/R_workshop" class="topic" style="color:red">Navigate Back</a>